---
title: "Part 2 - Working with Factors"
author: "Ted Laderas"
date: "5/18/2017"
output: 
  html_document:
    code_download: true
    code_folding: hide
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

In this section, we'll learn some skills about manipulating factor (categorical) data.

We'll do this by making a bar plot progressively making it more complex. We'll then dive into boxplots, a way to view the distribution of a variable across the categories of a different variable.

## Reviewing Factors 

Factors are frequently how R represents categorical data.

There are two kinds of factors: 

+ `factor` - used for *nominal* data ("Ducks", "Cats", "Dogs")`
+ `ordered`- used for *ordinal* data ("10-30","31-40","41-60")

We'll manipulate our barplot and add more information using factors.

Here's the simple dataset we'll use to investigate how to work with factors in `ggplot`. Here we also introduce how to read in data stored in an R data file (an RDS) into a data frame in R. This also acts as a primer to when we discuss importing data in the afternoon workshop. 

We display the `pets` data frame by simply running the line containing `pets`.

```{r}
library(tidyverse)
pets <- read_rds("data/pets.rds")
pets
```

We see that `name` and `animal` are stored as `factor`s and `age_category` is stored as an ordinal `factor`. Note that `name` has three levels ordered as `"0-2"`, `"3-4"`, and `"5-8"`.

## A Basic Barplot using `geom_bar()`

By default, the `geom_bar()` function counts the number of values with each factor level. Note that you don't map to a y-aesthetic here, because the y values are the counts.

Given this dataset, we might want to ask how many pets have the same name. What is the most popular name?

```{r}
ggplot(data = pets, mapping = aes(x = name)) + 
  geom_bar()
```

## Your Task

Try mapping another variable to `fill` (try both `weight` and `animal`). Are the results what you expected?

```{r}
# Test out `weight` and `animal` here



```


## Faceting

Say you have another `factor` variable and you want to stratify the plots based on that. You can do that by supplying the name of that variable as a facet. Here, we facet our barplot by `shots_current`.

```{r}
ggplot(data = pets, mapping = aes(x = name)) + 
  geom_bar() + 
  ## have to specify facets using notation
  facet_wrap(facets =~ shots_current) + 
  ## we make the x axis x angled for better legibility
  theme(axis.text.x = element_text(angle=45))
```

You might notice that there are blank spots for the categories in each facet. We can restrict the factors in each by using `scale = "free_x"` argument in `facet_wrap()`.

## Your Task

How many animals named "Morris" did not receive shots?

```{r}

```


What happens when you replace the `scale` argument with "free_y"?

```{r}
ggplot(pets, aes(x = name)) + geom_bar() + 
  facet_wrap(facets =~ shots_current, scale = "free_x") +
  theme(axis.text.x = element_text(angle = 45))
```

## Stacked Bars

Let's see how many of each animal got shots. We can do this by mapping `shots_current` to `fill`.

```{r}
#we map color (the outline of the plot) to black to make it look prettier
ggplot(pets, aes(x = animal, fill = shots_current)) + 
  geom_bar(color = "black")
```

## Proportional Barchart

We may only be interested in the relative proportions between the different categories. Visualizing this is useful for various 2 x 2 tests on proportions.

What percent of dogs did not receive shots?

```{r}
ggplot(pets, aes(x = animal,fill = shots_current)) + 
  geom_bar(position = "fill", color = "black")
```

## Dodge those bars!

Instead of stacking, we can also dodge the bars (move the bars so they're beside each other).

```{r}
ggplot(pets, aes(x = animal, fill = shots_current)) + 
  geom_bar(position = "dodge", color = "black")
```

## Your Task: Bar Charts

Given the `pets` `data.frame`, plot a stacked proportional barchart that shows the age category counts by animal type. Is the proportion of animals receiving shots the same across each age category?

Hints: think about what to map to `x`, and what to map to `fill`.

Intermediate Folks: facet this plot by `shots_current`. 

```{r}
# Space for your answer here.

```

```{r}
# Intermediate: Space for your answer here.

```


## Boxplots

Boxplots allow us to assess distributions of a continuous variable conditioned on categorical variables.

What does this tell us? 

```{r}
ggplot(pets, aes(x = shots_current, y = weight)) + 
  geom_boxplot()
```

## Violin Plots

Violin plots are another useful way to visualize the data. They provide a more nuanced look at the data. They're a density plot that's mirrored around the y-axis.

```{r}
ggplot(pets, aes(x = age_category, y = weight, fill = age_category)) + 
  geom_violin()
```

## Your task: How heavy are our pets?

Visualize `weight` by `animal` type as both a boxplot and a violin plot. What do you conclude? Which kind of animal weighs more on average than the other?

Intermediate challenge: How would we plot both boxplots and a violin plot on the same graph?

```{r}
## Space for your answer here

```

```{r}
## Intermediate: Space for your answer here

```


## What you learned in this section

- Visualizing factor data
- Simple, stacked, stacked proportional, and dodged barplots 
- Faceting a graph
- Boxplots and violin plots